The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-28 rational.tex
hopefully undid some damage
abstract and intro

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}
\usepackage{bm}

@@

\begin{abstract}
  Every functional programmer knows about sum and product types, {{a+b}} and
  {{a*b}} respectively. Negative and fractional types, $a-b$ and $a/b$
  respectively, are much less known and their computational interpretation is
  unfamiliar and often complicated. We show that in a programming model in
  which information is preserved (such as the model introduced in our recent
  paper on \emph{Information Effects}), these types have particularly
  intuitive and natural computational interpretations. Intuitively, values of
  negative types are values that flow ``backwards'' to satisfy demands and
  values of fractional types are values that impose constraints on their
  context.  The combination of these negative and fractional types enables
  greater flexibility in programming by breaking global invariants into local
  ones that can be autonomously satisfied by a subcomputation. Furthermore,
  they allow a programmer to express a wide range of programming idioms,
  including higher-order functions, delimited continuations, speculative
  computation, logic programming, and more.
\end{abstract}

@@

@@

@@

@@

>>>>

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}
\usepackage{bm}
\usepackage{cmll}

@@

\begin{abstract}
  Every functional programmer knows about sum and product types, {{a+b}} and
  {{a*b}}. Negative and fractional types, $a-b$ and $a/b$ respectively, are
  much less known and their computational interpretation is unfamiliar and
  often complicated. We show that in a programming model in which information
  is preserved (such as the model introduced in our recent paper on
  \emph{Information Effects}), these types have particularly natural
  computational interpretations. Intuitively, values of negative types are
  values that flow ``backwards'' to satisfy demands and values of fractional
  types are values that impose constraints on their context.  The combination
  of these negative and fractional types enables greater flexibility in
  programming by breaking global invariants into local ones that can be
  autonomously satisfied by a subcomputation. Furthermore, these types and
  their properties suggest that the previously observed duality of
  computation conflated two orthogonal notions: an additive duality that
  corresponds to backtracking and a multiplicative duality that corresponds
  to unification.
\end{abstract}

@@

In 1989, Filinski~\cite{Filinski:1989:DCI:648332.755574} observed that values
and continuations are dual notions. This observation was followed by numerous
papers~\cite{Griffin:1989:FNC:96709.96714, curien00duality, Wadler:2003,
  DBLP:conf/rta/Wadler05} which explored and refined this duality. Even
earlier, Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2}
introduced a logic which contains a dual to implication. Her work has been
distilled in the form of \emph{subtractive logic}~\cite{Crolard01} which has
recently been related to coroutines~\cite{Crolard01082004} and delimited
continuations~\cite{Ariola:2009:TFD:1743339.1743381}.

In a seemingly unrelated development, Girard in 1987 introduced linear
logic~\cite{Girard87tcs} which, among other contributions, exposed an
additive/multiplicative distinction in logical connectives and rules. In
particular, linear logic includes additive disjunctions $\oplus$ and
conjunctions $\with$ as well as multiplicative disjunctions $\parr$ and
conjuctions $\otimes$. Duality is also prominent in linear logic: it relates
the additive connectives to each other (the dual of $\oplus$ is $\with$ and
vice-versa) and the multiplicative connectives to each other (the dual of
$\otimes$ is $\parr$ and vice-versa).

@@

\begin{multicols}{2}
\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eps_plus1.pdf}
}
\end{center}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eps_plus2.pdf}
}
\end{center}  
\end{multicols}

@@

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/eta_times.pdf}
\end{center}

\begin{center}
  \includegraphics{diagrams/eps_times.pdf}
\end{center}
\end{multicols}

@@

test test test
\begin{multicols}{2}
\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eta_times1.pdf}
}
\end{center}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eps_times1.pdf}
}
\end{center}  
\end{multicols}
